Northern Pikeminnow Abundance in Deadwater Slough, Salmon River, Idaho, and Potential Impacts to Local Chinook Salmon Populations

### 2022-04-07 analysis/paper/paper.Rmd
near final edits to Methods

The Deadwater Slough is an approximately `r round(ddw_rkms, 1)` km section of the mainstem Salmon River located roughly 6 river kilometers (rkm) downstream from the town of North Fork, Idaho (Figure \@ref(fig:map)). The downstream end of the slough is located at the confluence of Dump Creek and the Salmon River. Around 1897, the failure of a small mining diversion reservoir in the Dump Creek drainage resulted in an erosion event and upslope instability that deposited substantial amounts of sediment at its confluence with the Salmon River [@Emerson1973]. That sediment formed an alluvial fan which created a slow, deep section in the river, spanning approximately `r round(ddw_hectares)` hectares, averaging `r round(ddw_avg_width_m)` m wide, and resembling a small reservoir.

@@

We estimated the population size of piscine predators in the Deadwater Slough using a mark-recapture survey design and a catch per unit effort (CPUE) approach. Predators were sampled near the peaks of the fall DSR and spring NRR emigrations. During our initial survey in 2019, Northern Pikeminnow were the most prevalent piscine predator in Deadwater Slough and consequently became our focal taxa. Our intent was to survey during the emigration of the 2018 Chinook Salmon brood year that occurred fall 2019 and spring 2020, however logistical constraints during the onset of the Covid-19 pandemic delayed the spring 2020 survey until 2021. Over this period, an additional fall survey occurred during the 2020 DSR emigration. Fall surveys were constrained to two weeks to minimize Northern Pikeminnow immigration/emigration.

@@

where $M$, $n$, and $m$ are indexed by each survey (day), $i$. The 95% confidence intervals were calculated using a Poisson approximation [@Krebs1999].<!--The Schnabel estimator is essentially the weighted average of a series of Chapman-modified Lincoln-Peterson estimators.-->

Because fish capture was dependent on volunteer anglers, there was concern that sampling effort was biased to locations preferred by anglers. If release sites for marked fish were not proximal to preferred angling locales, recently released fish may have been less susceptible to immediate capture than unmarked fish. To account for this potential bias, we included a "Delayed-mixing" Schnabel estimator that assumed fish marked and released on a given survey day were not available for recapture for 48 hours. The "Delayed-mixing" Schnabel estimator, is given by:

$$
\hat{N} = \frac{\sum\limits_{i = 1}^k n_i M_{i-1}}{\left(\sum\limits_{i = 1}^k m_i \right) + 1}
$$
All mark-recapture abundance estimators assumed: (1) the population is closed (no immigration, emigration, births or deaths during the survey period), (2) all fish have equal chance of being caught in subsequent survey days, (3) marking a fish does not affect its chances of recapture, (4) no loss of marks, and (5) marks are not missed or misidentified.

@@

To estimate the total consumption potential of Northern Pikeminnow in Deadwater Slough during the peaks of fall DSR and spring NRR emigrations, we used the Fish Bioenergetics v4.0 application developed by @Deslauriers2017 applied in R statistical software [@RCoreTeam2021]. The daily rate of consumption in grams for an individual Northern Pikeminnow was estimated based on predator and prey energy densities, predator start and end weights, and water temperatures. Separate models were run for the fall period (September 15 - November 30) and the spring period (March 1 - May 31) to coincide with peak emigrations of DSR and NRR juveniles from the Lemhi River, the largest Chinook Salmon population in the Upper Salmon MPG.

@@

This resulted in average fall and spring starting weights of `r round(avg_fall_NPM_g, 1)` and `r round(avg_spring_NPM_g, 1)` g for Northern Pikeminnow with average TLs of `r round(avg_fall_NPM_TL, 1)` and `r round(avg_spring_NPM_TL, 1)` mm, respectively. We assumed no growth in individual Northern Pikeminnow, resulting in equal start and end weights. Mean daily water temperatures were summarized from 15-minute interval temperature readings between March 3, 2013 and June 14, 2021 from USGS gage station 13307000, approximately 22 rkm downstream of Deadwater Slough.

The proportion of fish in the Northern Pikeminnow's diet relative to non-fish prey items is unknown in Deadwater Slough. Therefore, we conducted a series of model runs with varying proportions of fish in the diet ranging from 30 - 90%, in 10% increments. These values were supported by observations elsewhere in the Columbia River basin (54-86%: @Shively1996; 48-86%: @Zimmerman1999; 37%: @Gray2001) that suggest fish are the majority of Northern Pikeminnow prey by volume. A model run was conducted for each combination of diet scenario and season, resulting in fourteen estimates of the total grams of fish consumed by an individual Northern Pikeminnow (Table \@ref(tab:bio-tbl)). To estimate the total biomass of fish consumed by Northern Pikeminnow in Deadwater Slough, we multiplied the grams of fish consumed by an individual Northern Pikeminnow by the estimated Northern Pikeminnow population sizes during fall and spring.

@@

We obtained an estimate of the number of juvenile Chinook Salmon consumed by dividing total biomass consumed by the average weight of DSR (`r avg_DSR_g` g) and NRR (`r avg_NRR_g` g) emigrants in the Upper Salmon MPG. The average weights of DSR and NRR emigrants were calculated from fish captured at seven rotary screw traps upstream of Deadwater Slough during the fall and spring periods. We also quantified the potential impact of Northern Pikeminnow predation on juvenile Chinook Salmon to adult returns by estimating the number of additional adult Chinook Salmon expected to return to Lower Granite Dam if predation by Pikeminnow was reduced or eliminated in the Deadwater Slough. We multiplied the estimated total juvenile Chinook Salmon consumed by the median Granite-to-Granite smolt-to-adult return rate (SAR) of 0.00614 (SD = 0.00051) from @McCann2019 for Upper Salmon River MPG Chinook Salmon.

@@

All data and code for the analyses presented here can be found in a GitHub repository at ***zenodo DOI*** here.

>>>>

The Deadwater Slough is an approximately `r round(ddw_rkms, 1)` km section of the mainstem Salmon River located roughly 6 river kilometers (rkm) downstream from the town of North Fork, Idaho (Figure \@ref(fig:map)). The downstream end of the slough is located at the confluence of Dump Creek and the Salmon River. A large alluvial fan at the mouth of Dump Creek has, at least partially, created a hydraulic control in the Salmon River resulting in the formation of Deadwater Slough. Although the origin and timing of the alluvial fan and Deadwater Slough is somewhat ambiguous [@Reichmuth1958; @USACE1986], they likely coincided with the failure of a small mining reservoir in the Dump Creek drainage around 1897 that deposited substantial amounts of sediment at its confluence with the Salmon River [@Emerson1973]. The slough is currently a slow, deep section in the river, spanning approximately `r round(ddw_hectares)` hectares, averaging `r round(ddw_avg_width_m)` m wide, and has characteristics resembling a small reservoir.

@@

We estimated the population size of piscine predators in the Deadwater Slough using a mark-recapture survey design and a catch per unit effort (CPUE) approach. Predators were sampled near the peaks of the fall DSR and spring NRR emigrations. During our initial survey in 2019, Northern Pikeminnow were the most prevalent piscine predator in Deadwater Slough and consequently became our focal taxa. Our intent was to survey during the emigration of the 2018 Chinook Salmon brood year that occurred fall 2019 and spring 2020, however logistical constraints during the onset of the Covid-19 pandemic delayed the spring 2020 survey until 2021. Over this period, an additional fall survey occurred during the 2020 DSR emigration. Fall surveys were constrained to two weeks to minimize Northern Pikeminnow immigration and emigration.

@@

where $M$, $n$, and $m$ are indexed by each survey (day), $i$. The 95% confidence intervals of the Schnabel estimator were calculated using a Poisson approximation [@Krebs1999].<!--The Schnabel estimator is essentially the weighted average of a series of Chapman-modified Lincoln-Peterson estimators.-->

Because fish capture was dependent on volunteer anglers, there was concern that sampling effort was biased to locations preferred by anglers. If release sites for marked fish were not proximal to preferred angling locales, recently released fish may have been less susceptible to immediate capture than unmarked fish. To account for this potential bias, we included an adjusted "delayed-mixing" Schnabel estimator that assumed fish marked and released on a given survey day were not available for recapture for 48 hours. The delayed-mixing Schnabel estimator, is given by:

$$
\hat{N} = \frac{\sum\limits_{i = 1}^k n_i M_{i-1}}{\left(\sum\limits_{i = 1}^k m_i \right) + 1}
$$
All mark-recapture abundance estimators include a Chapman correction for small sample sizes [@Chapman1951] and assumed: (1) the population is closed (no immigration, emigration, births or deaths during the survey period), (2) all fish have equal chance of being caught in subsequent survey days, (3) marking a fish does not affect its chances of recapture, (4) no loss of marks, and (5) marks are not missed or misidentified.

@@

To estimate the total consumption potential of Northern Pikeminnow in Deadwater Slough during the peaks of fall DSR and spring NRR emigrations, we used the Fish Bioenergetics v4.0 application developed by @Deslauriers2017 applied in R statistical software [@RCoreTeam2021]. The daily rate of consumption in grams for an individual Northern Pikeminnow was estimated based on predator and prey energy densities, predator start and end weights, and water temperatures. Separate models were run for the fall period (September 15 - November 30) and the spring period (March 1 - May 31) to coincide with peak emigrations of DSR and NRR juveniles from the Lemhi River, the largest Chinook Salmon population in the Upper Salmon River.

@@

This resulted in average fall DSR and spring NRR starting weights of `r round(avg_fall_NPM_g, 1)` and `r round(avg_spring_NPM_g, 1)` g for Northern Pikeminnow with average TLs of `r round(avg_fall_NPM_TL, 1)` and `r round(avg_spring_NPM_TL, 1)` mm, respectively. We assumed no growth in individual Northern Pikeminnow, resulting in equal start and end weights. Mean daily water temperatures were summarized from 15-minute interval temperature readings between March 3, 2013 and June 14, 2021 from USGS gage station 13307000, approximately 22 rkm downstream of Deadwater Slough.

The proportion of fish in the Northern Pikeminnow's diet relative to non-fish prey items is unknown in Deadwater Slough. Therefore, we conducted a series of model runs with varying proportions of fish in the diet ranging from 30 - 90%, in 10% increments. These values were supported by observations elsewhere in the Columbia River basin (54-86%: @Shively1996; 48-86%: @Zimmerman1999; 37%: @Gray2001) that suggest fish are the majority of Northern Pikeminnow prey by volume. A model run was conducted for each combination of diet scenario and season, resulting in fourteen estimates of the total grams of fish consumed by an individual Northern Pikeminnow. To estimate the total biomass of fish consumed by Northern Pikeminnow in Deadwater Slough, we multiplied the grams of fish consumed by an individual Northern Pikeminnow by the estimated Northern Pikeminnow population sizes during fall and spring.

@@

We obtained an estimate of the number of juvenile Chinook Salmon consumed by dividing total biomass consumed by the average weight of DSR (`r avg_DSR_g` g) and NRR (`r avg_NRR_g` g) emigrants in the Upper Salmon MPG. The average weights of DSR and NRR emigrants were calculated from fish captured at seven rotary screw traps upstream of Deadwater Slough during the fall and spring periods. Although the primary impact to Chinook Salmon in Deadwater Slough by Northern Pikeminnow is juvenile predation, the common metric to evaluate salmon recovery is adult returns. We quantified the potential impact of Northern Pikeminnow predation on adult returns by estimating the number of equivalent adults expected to return to Lower Granite Dam if predation by Pikeminnow was eliminated in the Deadwater Slough. To accomplish this, we multiplied the estimated total juvenile Chinook Salmon consumed by the median Granite-to-Granite smolt-to-adult return rate (SAR) of 0.00614 (SD = 0.00051) from @McCann2019 for Chinook Salmon in the Upper Salmon River.

@@

All data and code for the analyses presented here can be found in a GitHub repository at ***zenodo DOI***.
